---
title: "Cluster Analysis for Crime Data"
author: "Clustering INSOFE Lab Assignment"
date: "16 July 2017"
output:
  html_document:
    toc: yes
    toc_depth: 3
    toc_float: yes
---

**NOTE** Before starting this assignment please remember to clear your environment, you can do that by running the following code chunk

```{r}
rm(list=ls(all=TRUE))

```

# Agenda

* Get the data

* Data pre-processing

* Explore the data

* Hierarchical Clustering

* Kmeans Clustering

* Visualising Clusters and Evaluation


# Problem Description

* In the following Unsupervised Learning activity, you will perform cluster analysis on a dataset that has arrests per 100,000 residents for assault, murder, and rape in each of the 50 US states

* The variable names in the dataset are self explanatory

* So, you will cluster US states based on the crime rates, which can then be passed on to public policy experts to gain insight from it


# Reading & Understanding the Data

* Read in the dataset

```{r}

# Use the setwd() function to get to the directory where the data is present

setwd("C:\\insofe\\day13\\20170716_Batch30_CSE7305c_ClusterAnalysis_FullLab")

cluster_data=read.csv("crime_data.csv")

```

* Use the str() and summary() functions to get a feel for the dataset.

```{r}
str(cluster_data)
summary(cluster_data)


```

* Take a look at the data using the "head()" and "tail()" functions

```{r}

head(cluster_data)
tail(cluster_data)

```

# Data pre-processing

* Check for the number of missing values in the dataset

```{r}


sum(is.na(cluster_data))

```


* Convert the State names into row names and remove that variable from the dataset

```{r}

rownames(cluster_data)=cluster_data$State
cluster_data_rm <- cluster_data[, -c(colnames(cluster_data) %in% ("State"))]


```

* Standardize and scale the data
```{r}

cluster_data_scaled <- scale(cluster_data_rm, center = T, scale = T)

```

# Data exploration

* Visualise the distances between the individual observations using the fviz_dist()

```{r, fig.width=12, fig.height=8}
library(factoextra)
distance <- get_dist(cluster_data_scaled)
fviz_dist(distance, gradient = list(low = "#00AFBB", mid = "white", high = "#FC4E07"))
```

# Hierarchical Clustering

* Cluster the data using the Ward algorithm

```{r}


# We use the euclidean distance measure

dist <- dist(cluster_data_scaled, method = "euclidean")

hc_fit <- hclust(dist, method = "ward.D2")

```

* Plot the dendogram for hierarchical clustering

```{r, fig.height=5, fig.width=10}

plot(hc_fit)

```

* Cut the tree to 4 clusters

```{r}

points_hc <- cutree(hc_fit, k = 4)

# Store the clusters in a data frame along with the crime data

Crime_clust_hc <- cbind(points_hc, cluster_data_scaled)

# Have a look at the head of the new data frame

colnames(Crime_clust_hc)[1] <- "cluster_hc"

head(Crime_clust_hc)

```

* Plot a new dendogram, with each of the clusters being surrounded by a border, using the rect.hclust() function

```{r, fig.height=5, fig.width=10}

plot(hc_fit)
rect.hclust(hc_fit, k = 4, border = "red")

```


# K-Means Clustering

* Build a basic kmeans model with k = 2

```{r}

set.seed(123)
km_basic <- kmeans(cluster_data_scaled, centers = 2)

str(km_basic)
fviz_cluster(km_basic, cluster_data_scaled)

```

* Build a scree plot and choose a "k"

```{r}


# Initialize wss to 0

wss <- 0

# From 1 upto upto 10 cluster centers, fit the kmeans model

for (i in 1:10) {
  
  cfit = kmeans(cluster_data_scaled, centers = i)
  
  # Store the sum of within sum of square
  
  wss[i] <- sum(cfit$withinss)
  
}
plot(1:10, wss, type = "b")


set.seed(123)

fviz_nbclust(cluster_data_scaled, kmeans, method = "wss")

```

* Choose a k and cluster the data

```{r}

# We can choose a k = 4, 5 or 6


km_clust <- kmeans(cluster_data_scaled, 6, nstart = 10)

# after choosing k as 6, let's store the cluster groupings along with the data in a new data frame

km_points <- km_clust$cluster

# Store the cluster assignments in a new data frame

crime_clusts_km <- as.data.frame(cbind(km_clust$cluster, cluster_data_scaled))

# Look at the head of the data

head(crime_clusts_km)

colnames(crime_clusts_km)[1] <- "cluster_km"

```

* Visualise the clusters by plotting the data points on the first two principal components

```{r, fig.height=5, fig.width=8}

fviz_cluster(km_clust, cluster_data_scaled)

```

